\documentclass{article}

\usepackage{microtype}

\usepackage{amsmath}
\usepackage{mathtools}

\title{CS225 Spring 2018---Final Project Proposal}
\author
{
  William Slocum \\ \small{\texttt{@WilliamSlocum}} 
  \\ \small{\texttt{https://github.com/WilliamSlocum/CS225-FINAL}}
}
\date{\today}

\begin{document}
\maketitle

\section*{Project: Type Reconstruction}

We propose to implement Type Reconstruction, which will support powerful Type Inference, reducing the need for explicit Type Annotations.

\paragraph{Base Language}

We will work with a Language containing Booleans, Natural Numbers, Functions, and Let-Bindings.

\paragraph{Constraint-Based Typing} Without explicit Type Annotations, it isn't always possible to tell if an Expression is well-typed. Instead, we will calculate a set of possibilities where such an Expression would be well-typed. First, some definitions...\\
\\
Let $\Gamma$ be a context and $t$ a term. A solution for ($\Gamma$, $t$) is a pair ($\sigma$, $T$) such that $\sigma$ $\Gamma$ $\vdash$ $\sigma$ $t$ : $T$.\\
\\
A Constraint Set $C$ is a set of equations $\{S_{i} = T_{i}\}$. \\
\\
We say a substitution $\sigma$ unifies a set $C$ if for every equation $S = T$ in $C$, the substitution instances of $\sigma S$ and $\sigma T$ are identical. \\
\\
We will implement Constraint Based Typing by writing an algorithm that, given a context $\Gamma$ and a term $t$, finds a set of constraints.\\
\\
Then, we will use a Unification Algorithm to calculate Solutions to those Constraints.

\paragraph{Project Goals}

For this Project, I hope to complete...
\begin{enumerate}
\item A Constraint-Based Typing Function
\item A Unification Functions that calculates solutions to Constraint Sets
\end{enumerate}

\paragraph{Expected Challenges}

I anticipate the beginning of this Project will be the most challenging; structuring the OCaml code, 
understanding the various concepts. Once I get going, I hope things will run smoothly.

\paragraph{Timeline and Milestones} \noindent
By the Checkpoint I hope to have completed...
\begin{enumerate}
\item A Prototype Implementation of the Typing Function
\item A Prototype Implementation of the Unification Function
\item Several Test Cases for these Functions
\end{enumerate}

\noindent
By the Final Project Draft I hope to have completed...
\begin{enumerate}
\item Full Implementations of Typing and Unification Functions
\item A Comprehensive Test Suite
\item A Draft of a Presentation
\end{enumerate}

\noindent
By the Final Project Submission I hope to have completed...
\begin{enumerate}
\item A Finished Presentation
\item A Short Writeup
\item Any remaining Work that was Missing in the Draft
\end{enumerate}

\end{document}
